---
title: "EIF2A_proteomics"
author: "Howard Cen"
date: "2023-10-12"
output: html_document
---

```{r libraries}
# Install most recent version of ComplexHeatmap from Github
#install.packages("devtools") 
#library(devtools)
#install_github("jokergoo/ComplexHeatmap")
#install.packages("gridtext")
#install.packages("scales")


library(tidyverse) # for data wrangling and more
#library(readxl) # for reading excel file
library(ComplexHeatmap)
library(circlize) # installed with ComplexHeatmap, but need to load it
library(gridtext) # needed for complex heatmap
library(scales) # for calculating z-scores etc


setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) ##Set working directory

#dir.create("input")
#dir.create("output")
#dir.create("figures")

```




```{r heatmap}

library(RColorBrewer) 

df <- read.csv(file = "output/EIF2A_proteomics_annotated.csv")
View(df)

de <- df %>% filter(Welch.s.T.test.p.value.DMSO_OE_DMSO<0.005 |
                      Welch.s.T.test.p.value.TG_OE_TG<0.005 |
                      Welch.s.T.test.p.value.TG_DMSO<0.005 |
                      Welch.s.T.test.p.value.TG_OE_DMSO_OE<0.005)

de <- df %>% filter(Welch.s.T.test.p.value.DMSO_OE_DMSO<0.005 |
                      Welch.s.T.test.p.value.TG_OE_TG<0.005 )


de <- df %>% filter(Welch.s.T.test.q.value.TG_DMSO<0.05 |
                      Welch.s.T.test.q.value.TG_OE_DMSO_OE<0.05 )


de <- df %>% filter(Welch.s.T.test.p.value.TG_DMSO<0.005 |
                      Welch.s.T.test.p.value.TG_OE_TG<0.005)

de <- df %>% filter(Welch.s.T.test.p.value.TG_DMSO<0.005 |
                      Welch.s.T.test.p.value.TG_OE_DMSO_OE<0.005)
dim(de)
View(de)

m <- df[,4:15] 
rownames(m) <- df$Genes # set gene names as row names of m

m <- de[,4:15] # select certain columns or rows by df[rows,columns]
View(m)
rownames(m) <- de$Genes # set gene names as row names of m

m.z <- t(scale(t(m))) # scale -- calculate Z-score for columns: [(x-mean)/SD]; t -- transpose the matrix; log2 -- log2 transform the data
colnames(m.z)
max(m.z)
min(m.z)
View(m.z)

#rownames(m.z) <- rownames(m) # set gene names as rownames of m.z

# plot heatmap  --------------
# (originally established by Yi Han Xia)

number_of_dc <- length(grep("DMSO.ctrl", colnames(m.z))) # grep() -- find the position/index of, in this case, column names that contain the pattern "DN"
number_of_de <- length(grep("DMSO.EIF2A", colnames(m.z))) # length() -- get the length of things
number_of_tc <- length(grep("Tg.ctrl", colnames(m.z)))
number_of_te <- length(grep("Tg.EIF2A", colnames(m.z)))

number_of_dc
number_of_te

end_index_dc <- grep("DMSO.ctrl", colnames(m.z))[number_of_dc] # get the end position/index of each group
end_index_de <- grep("DMSO.EIF2A", colnames(m.z))[number_of_de]
end_index_tc <- grep("Tg.ctrl", colnames(m.z))[number_of_tc]
end_index_te <- grep("Tg.EIF2A", colnames(m.z))[number_of_te]

end_index_dc
end_index_te
#set.seed(12) # run this to get the same k-means clusters 
cbPalette <- c("#E69F00", #lightorange
               
                "#56B4E9", #blue
               "#009E73", #green
               
                "#CC79A7", #magenta
               "#D55E00", #darkorange
               "#999999", #grey
                
                
                "#0072B2", #darkblue
                "#CC79A7", #magenta
                
                "#F0E442" #yellow
               )


# heatmap for DMSO_EIF2A vs DMSO or Tg_EIF2A vs Tg                
heatmap <- Heatmap(m.z,
                   show_row_names = TRUE,
                   show_column_names = FALSE,
                   show_row_dend = TRUE,
                   show_column_dend = FALSE,
                   
                   clustering_distance_rows = "euclidean",
                   clustering_method_rows = "ward.D2", # default linkage method is "complete"
                   
                   row_dend_side = "left",
                   row_dend_width = unit(15, "mm"),
                   
                   column_names_side = "top",
                   column_dend_side = "bottom",
                   
                   row_labels = gt_render(rownames(m.z)), ### now we set row names (genes) as labels
                   row_names_gp = gpar(fontsize = 10),
                  
                   col = colorRamp2(c(-3,0,3), c("blue", "white", "red")), # set the color range from Z-score -3 to 3 (saturate the outliers,  if you didn't remove them).
                   #col = colorRamp2(c(min(na.omit(m.z)),0,max(na.omit(m.z))), c("blue", "white", "red")), # alternatively, set the range from min to max value
                   
                   column_order = 1:ncol(m.z),
                   
                   height = unit(220, "mm"), # set the height of heatmap
                   width = ncol(m.z)*unit(8, "mm"), # set the column width
                   
                   top_annotation = columnAnnotation(empty = anno_empty(border = FALSE, 
                                                                        height = unit(6, "mm"))), # leave an empty space on top for the condition labels below.
                   
                   right_annotation = 
                           rowAnnotation(`DMSO_EIF2A vs DMSO` = de$sig_DMSO_EIF2A.ctrl,
                                         `Tg_EIF2A vs Tg` = de$sig_Tg_EIF2A.ctrl,
                                         
                                         
                                         col = list(`DMSO_EIF2A vs DMSO` = c("TRUE" = cbPalette[1]),
                                                    `Tg_EIF2A vs Tg` = c("TRUE" = cbPalette[2])
                                         )
                                         ,
                                         annotation_legend_param = list(
                      
                                           `DMSO_EIF2A vs DMSO` = list(
                                             title = c(""),
                                             labels = "DMSO_EIF2A vs DMSO"#expression(italic("Ins1")^italic("+/+")*" Sucrose vs Water") #expression() issue - https://github.com/jokergoo/ComplexHeatmap/issues/678
                                           ),
                      
                                           `Tg_EIF2A vs Tg` = list(
                                             title = c(""),
                                             labels = "Tg_EIF2A vs Tg"
                                           )
                                         ),
                                         show_annotation_name = FALSE,
                                         show_legend=TRUE,
                                         na_col = "white",
                                         width = unit(8, "mm")
                      
                           ),

                   
                   
                   border_gp = gpar(col = "black"), # set border color
                   
                   show_heatmap_legend = TRUE,
                   heatmap_legend_param = list(
                     title = "Row z-score",
                     title_position = "topleft",
                     #at = c(-3, -2, -1, 0, 1, 2, 3), # set ticks/numbers of legend
                     legend_height = unit(4, "cm")),
                   
                   #split = 4, # split the heatmap into separate blocks by hierarchical clustering
                   #row_km = 4, # or split the heatmap into separate blocks by k-means clustering
                   ## determine the number of clusters: https://www.datanovia.com/en/lessons/determining-the-optimal-number-of-clusters-3-must-know-methods/
                   
                  layer_fun = function(j, i, x, y, width, height, fill) {
                     mat = restore_matrix(j, i, x, y)
                     
                     ind = unique(c(mat[, c(end_index_dc,
                                            end_index_de,
                                            end_index_tc
                                            ) # enter the number where you want a gap between columns
                                        ])) 
                     
                     grid.rect(x = x[ind] + unit(0.5/ncol(m.z), "npc"), 
                               y = y[ind], 
                               width = unit(0.03, "inches"), # width of the gap
                               height = unit(1/nrow(m.z), "npc"),
                               gp = gpar(col = "white", fill = "white") # color of the gap
                     )
                   }
                   ) 

draw(heatmap)





# heatmap for Tg_EIF2A vs Tg or Tg vs DMSO              
heatmap <- Heatmap(m.z,
                   show_row_names = TRUE,
                   show_column_names = FALSE,
                   show_row_dend = TRUE,
                   show_column_dend = FALSE,
                   
                   clustering_distance_rows = "euclidean",
                   clustering_method_rows = "ward.D2", # default linkage method is "complete"
                   
                   row_dend_side = "left",
                   row_dend_width = unit(15, "mm"),
                   
                   column_names_side = "top",
                   column_dend_side = "bottom",
                   
                   row_labels = gt_render(rownames(m.z)), ### now we set row names (genes) as labels
                   row_names_gp = gpar(fontsize = 10),
                  
                   col = colorRamp2(c(-3,0,3), c("blue", "white", "red")), # set the color range from Z-score -3 to 3 (saturate the outliers,  if you didn't remove them).
                   #col = colorRamp2(c(min(na.omit(m.z)),0,max(na.omit(m.z))), c("blue", "white", "red")), # alternatively, set the range from min to max value
                   
                   column_order = 1:ncol(m.z),
                   
                   height = unit(300, "mm"), # set the height of heatmap
                   width = ncol(m.z)*unit(8, "mm"), # set the column width
                   
                   top_annotation = columnAnnotation(empty = anno_empty(border = FALSE, 
                                                                        height = unit(6, "mm"))), # leave an empty space on top for the condition labels below.
                   
                   right_annotation = 
                           rowAnnotation(`Tg vs DMSO` = de$sig_Tg.DMSO_ctrl,
                                         `Tg_EIF2A vs Tg` = de$sig_Tg_EIF2A.ctrl,
                                         
                                         
                                         col = list(`Tg vs DMSO` = c("TRUE" = cbPalette[3]),
                                                    `Tg_EIF2A vs Tg` = c("TRUE" = cbPalette[2])
                                         )
                                         ,
                                         annotation_legend_param = list(
                      
                                           `Tg vs DMSO` = list(
                                             title = c(""),
                                             labels = "Tg vs DMSO"#expression(italic("Ins1")^italic("+/+")*" Sucrose vs Water") #expression() issue - https://github.com/jokergoo/ComplexHeatmap/issues/678
                                           ),
                      
                                           `Tg_EIF2A vs Tg` = list(
                                             title = c(""),
                                             labels = "Tg_EIF2A vs Tg"
                                           )
                                         ),
                                         show_annotation_name = FALSE,
                                         show_legend=TRUE,
                                         na_col = "white",
                                         width = unit(8, "mm")
                      
                           ),

                   
                   
                   border_gp = gpar(col = "black"), # set border color
                   
                   show_heatmap_legend = TRUE,
                   heatmap_legend_param = list(
                     title = "Row z-score",
                     title_position = "topleft",
                     #at = c(-3, -2, -1, 0, 1, 2, 3), # set ticks/numbers of legend
                     legend_height = unit(4, "cm")),
                   
                   #split = 4, # split the heatmap into separate blocks by hierarchical clustering
                   #row_km = 4, # or split the heatmap into separate blocks by k-means clustering
                   ## determine the number of clusters: https://www.datanovia.com/en/lessons/determining-the-optimal-number-of-clusters-3-must-know-methods/
                   
                  layer_fun = function(j, i, x, y, width, height, fill) {
                     mat = restore_matrix(j, i, x, y)
                     
                     ind = unique(c(mat[, c(end_index_dc,
                                            end_index_de,
                                            end_index_tc
                                            ) # enter the number where you want a gap between columns
                                        ])) 
                     
                     grid.rect(x = x[ind] + unit(0.5/ncol(m.z), "npc"), 
                               y = y[ind], 
                               width = unit(0.03, "inches"), # width of the gap
                               height = unit(1/nrow(m.z), "npc"),
                               gp = gpar(col = "white", fill = "white") # color of the gap
                     )
                   }
                   ) 

draw(heatmap)





# heatmap for Tg_EIF2A vs DMSO_EIF2A or Tg vs DMSO              
heatmap <- Heatmap(m.z,
                   show_row_names = TRUE,
                   show_column_names = FALSE,
                   show_row_dend = TRUE,
                   show_column_dend = FALSE,
                   
                   clustering_distance_rows = "euclidean",
                   clustering_method_rows = "ward.D2", # default linkage method is "complete"
                   
                   row_dend_side = "left",
                   row_dend_width = unit(15, "mm"),
                   
                   column_names_side = "top",
                   column_dend_side = "bottom",
                   
                   row_labels = gt_render(rownames(m.z)), ### now we set row names (genes) as labels
                   row_names_gp = gpar(fontsize = 10),
                  
                   col = colorRamp2(c(-3,0,3), c("blue", "white", "red")), # set the color range from Z-score -3 to 3 (saturate the outliers,  if you didn't remove them).
                   #col = colorRamp2(c(min(na.omit(m.z)),0,max(na.omit(m.z))), c("blue", "white", "red")), # alternatively, set the range from min to max value
                   
                   column_order = 1:ncol(m.z),
                   
                   height = unit(600, "mm"), # set the height of heatmap
                   width = ncol(m.z)*unit(8, "mm"), # set the column width
                   
                   top_annotation = columnAnnotation(empty = anno_empty(border = FALSE, 
                                                                        height = unit(6, "mm"))), # leave an empty space on top for the condition labels below.
                   
                   right_annotation = 
                           rowAnnotation(`Tg vs DMSO` = de$sig_Tg.DMSO_ctrl,
                                         `Tg_EIF2A vs DMSO_EIF2A` = de$sig_Tg.DMSO_EIF2A,
                                         
                                         
                                         col = list(`Tg vs DMSO` = c("TRUE" = cbPalette[3]),
                                                    `Tg_EIF2A vs DMSO_EIF2A` = c("TRUE" = cbPalette[4])
                                         )
                                         ,
                                         annotation_legend_param = list(
                      
                                           `Tg vs DMSO` = list(
                                             title = c(""),
                                             labels = "Tg vs DMSO"#expression(italic("Ins1")^italic("+/+")*" Sucrose vs Water") #expression() issue - https://github.com/jokergoo/ComplexHeatmap/issues/678
                                           ),
                      
                                           `Tg_EIF2A vs DMSO_EIF2A` = list(
                                             title = c(""),
                                             labels = "Tg_EIF2A vs DMSO_EIF2A"
                                           )
                                         ),
                                         show_annotation_name = FALSE,
                                         show_legend=TRUE,
                                         na_col = "white",
                                         width = unit(8, "mm")
                      
                           ),

                   
                   
                   border_gp = gpar(col = "black"), # set border color
                   
                   show_heatmap_legend = TRUE,
                   heatmap_legend_param = list(
                     title = "Row z-score",
                     title_position = "topleft",
                     #at = c(-3, -2, -1, 0, 1, 2, 3), # set ticks/numbers of legend
                     legend_height = unit(4, "cm")),
                   
                   #split = 4, # split the heatmap into separate blocks by hierarchical clustering
                   #row_km = 4, # or split the heatmap into separate blocks by k-means clustering
                   ## determine the number of clusters: https://www.datanovia.com/en/lessons/determining-the-optimal-number-of-clusters-3-must-know-methods/
                   
                  layer_fun = function(j, i, x, y, width, height, fill) {
                     mat = restore_matrix(j, i, x, y)
                     
                     ind = unique(c(mat[, c(end_index_dc,
                                            end_index_de,
                                            end_index_tc
                                            ) # enter the number where you want a gap between columns
                                        ])) 
                     
                     grid.rect(x = x[ind] + unit(0.5/ncol(m.z), "npc"), 
                               y = y[ind], 
                               width = unit(0.03, "inches"), # width of the gap
                               height = unit(1/nrow(m.z), "npc"),
                               gp = gpar(col = "white", fill = "white") # color of the gap
                     )
                   }
                   ) 

draw(heatmap)
```
# save and label heatmap ---------------

#step 1 - save a pdf file
pdf(file = 'figures/heatmap_EIF2A_proteomics_DE.pdf',
    width = 8, 
    height = 10)
    
pdf(file = 'figures/heatmap_EIF2A_proteomics_DE_reverse.pdf',
    width = 8, 
    height = 13)
    
pdf(file = 'figures/heatmap_EIF2A_proteomics_DE_oe.pdf',
    width = 8, 
    height = 24)

#step 2 - draw heatmap in the saved file
draw(heatmap)

#step 3 - draw the labeling on the heatmap in the saved file

# labeling heatmap
#list_components() # get the viewport names
seekViewport("annotation_empty_1") # seek the empty space we saved at the top of heatmap, called "annotation_empty_1"

#Condition label 1

## draw a text label
grid.text("DMSO", 
          x = (end_index_dc)/2/ncol(m.z), # location at the middle of DN
          y = 0.5,
          just = c("center", "center"),
          gp = gpar(fontsize = 11,
                    col = "black"))

#Condition label 2

grid.text("DMSO_EIF2A", 
          x = (end_index_dc + end_index_de)/2/ncol(m.z), # location at the middle of DH
          y = 0.5,
          just = c("center", "center"),
          gp = gpar(fontsize = 11))

#Condition label 3


grid.text("Tg", 
          x = (end_index_de + end_index_tc)/2/ncol(m.z),
          y = 0.5,
          just = c("center", "center"),
          gp = gpar(fontsize = 11))

#Condition label 4


grid.text("Tg_EIF2A", 
          x = (end_index_tc + end_index_te)/2/ncol(m.z),
          y = 0.5,
          just = c("center", "center"),
          gp = gpar(fontsize = 11))


dev.off()

## Top label vertical gaps
grid.rect(x = end_index_DN/ncol(m.z),
          y = 0,
          height = 1,
          width = unit(0.03, "inches"), # same width as the width of gaps in heatmap
          just = c("center", "bottom"),
          gp = gpar(fill = "white", 
                    col = "white", 
                    lwd = 1
          ))
grid.rect(x = end_index_DH/ncol(m.z),
          y = 0,
          height = 1,
          width = unit(0.03, "inches"),
          just = c("center", "bottom"),
          gp = gpar(fill = "white", 
                    col = "white", 
                    lwd = 1
          ))
grid.rect(x = end_index_IN/ncol(m.z),
          y = 0,
          height = 1,
          width = unit(0.03, "inches"),
          just = c("center", "bottom"),
          gp = gpar(fill = "white", 
                    col = "white", 
                    lwd = 1
          ))

# step 4 - close the plotting device (the saved file)
dev.off()


```