---
title: "protti"
author: "Howard Cen"
date: "2024-04-05"
output: html_document
---

```{r }
# Author: Enes K. Ergin
# Date: 2023-07-13

## Load the packages required

# Install the packages if not installed
if (!require("protti", quietly = TRUE))
    devtools::install_github("jpquast/protti", dependencies = TRUE)

if (!requireNamespace("tidyverse", quietly = TRUE))
    install.packages("tidyverse")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
library(tidyverse) # Data manipulation
library(protti) # Main proteomics analysis package
# Load functions from utils.R
source("utils.R")
packageVersion("S4Arrays")
#remove.packages("S4Arrays")
#BiocManager::install("S4Arrays")
## Prepare the Data for Analysis


# Protein-Level Data
protein_data <- read_csv("output/EIF2A_proteomics_raw.csv")[,-c(1,2)]
protein_data <- protein_data[!is.na(protein_data$Genes),] # remove proteins with missing gene names
View(protein_data)
protein_data <- protein_data[rowSums(is.na(protein_data[-1])) < 0.5*ncol(protein_data[-1]), ]
length(protein_data$Genes) == length(unique(protein_data$Genes)) # TRUE, all gene names are unique


# meta data
metadata <- data.frame(ID = colnames(protein_data)[-1])
metadata <- separate(data = metadata, col = ID, into = c("Tg", "EIF2A", "Replica"), remove = F) %>%
  mutate(Group = paste0(Tg,"_",EIF2A))
View(metadata)
# Find maximum number of Replicates
max(metadata$Replica)

```


```{r}
# Convert the peptide_data to long format
protein_data_long <- tidyr::pivot_longer(
    protein_data,
    cols = -c("Genes"),
    names_to = "ID",
    values_to = "Intensity"
)
View(protein_data_long)

# Merge the metadata and peptide_data_longer
protein_data_long <- dplyr::left_join(
    protein_data_long,
    metadata,
    by = "ID"
)

# Rename the columns
protein_data_long <- dplyr::rename(
    protein_data_long,
    "Protein" = "Genes",
    "Sample" = "ID"
)

# Create a log2 transformed intensity column
protein_data_long$Intensity_log2 <- log2(protein_data_long$Intensity)

# BONUS 1 - Synthetic Data Generation
# Creates 100 Proteins and their replicates with 5 replicates over 2 conditions
#data <- protti::create_synthetic_data(
#    n_proteins = 100,       # number of proteins
#    frac_change = 0.05,
#    n_replicates = 5,
#    n_conditions = 2,
#    method = "effect_random",
#    additional_metadata = TRUE
#)


## Quality Checks and Data Filtering

# 1. Number of Identified Proteins per Samples
plot <- qc_ids(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    condition = EIF2A,
    intensity = Intensity,
    plot = TRUE
)

# Visualize the plot

# Sort the bars by the number of proteins
plot + 
  ggplot2::coord_flip() + 
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 0, hjust = 0.5))

# Rotate the x-axis labels
plot + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5)
)
```

```{r}

# 2. Intensity Distribution and Median Intensity Plots
#   a. Intensity Distribution
qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = Intensity,
    plot_style = "boxplot"
)

# Plot again with log2 transformed intensity
qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = Intensity_log2,
    plot_style = "boxplot"
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```
```{r}
#   b. Median Intensity Plot
qc_median_intensities(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r}

# 3. Co-efficients of Variation (CV)
# Within Sample CVs - Usually 10 or 20% are good
qc_cvs(
    data = protein_data_long,
    grouping = Protein,
    condition = Tg,
    intensity = Intensity,
    plot = TRUE,
    plot_style = "boxplot"
    #max_cv = 1000
)
?qc_cvs
```

```{r}

# Within SampleType CVs (Patient, Cell Line, PDX)
qc_cvs(
    data = protein_data_long,
    grouping = Protein,
    condition = EIF2A,
    intensity = Intensity,
    plot = TRUE,
    plot_style = "boxplot"
)
```


```{r}
# CVs between Samples in
# Patient+BALL, Patient+ALL,
# Cell Line+BALL, Cell Line+ALL
# PDX+BALL, PDX+ALL
# Combine CVs between SampleType+Disease
#protein_data_long$Group <- paste(
#    protein_data_long$Tg,
#    protein_data_long$EIF2A,
#    sep = "_"
#)

qc_cvs(
    data = protein_data_long,
    grouping = Protein,
    condition = Group,
    intensity = Intensity,
    plot = TRUE,
    plot_style = "boxplot"
)
```
```{r}
# 4. Data Completeness -  what percentage of all detected precursors is present in each sample.
qc_data_completeness(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2,
    plot = TRUE
)
?qc_data_completeness
```


```{r}
# 5. Sample Correlation
# NOTE: Requires pheatmap & seriation packages
#install.packages('seriation')

qc_sample_correlation(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2,
    condition = Group,
    interactive = FALSE,
    method = "pearson"
)
# can't display the heatmap when it's ran inside code chunk. Works outside below
```


qc_sample_correlation(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2,
    condition = Group,
    interactive = FALSE,
    method = "pearson"
)

```{r}
# 6. Principal Component Analysis (PCA)

qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = Intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "scree"
)
```

```{r}
qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = Intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "pca"
)
```
```{r}
# 7. Ranked Intensity Distribution (Protein-Rank Plot)
qc_ranked_intensities(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity_log2 = Intensity_log2,
  plot = TRUE,
  y_axis_transformation = "log2"
)
```
```{r}
# 8. Removing Problematic Samples
# Remove instances where SampleName == "NP21"
#protein_data_long <- dplyr::filter(
#    protein_data_long,
#    SampleName != "NP21"
#)

# 9. Removing Highly Missing Proteins
# Remove proteins with missing values in more than 75% of samples
# TODO: This is a custom solution, WIP


#colnames(protein_data_long)
#nsample <- length(unique(protein_data_long$Sample))
#protein_data_long_filter <- protein_data_long %>%
#  group_by(Protein) %>%
#  filter(sum(is.na(Intensity_log2))<nsample*0.75)

```



```{r}
## Data Normalization

# Median Normalization using protti
protein_data_long <- normalise(
    data = protein_data_long,
    sample = Sample,
    intensity = Intensity_log2,
    method = "median"
) # Adds normalized_"intensity" column

# Plot again with log2 transformed intensity
qc_intensity_distribution(
    data = protein_data_long_filter,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = normalised_intensity_log2,
    plot_style = "boxplot"
)
```

```{r}
b1 <- qc_median_intensities(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2
) + ggplot2::ggtitle("Before Normalization")

b2 <- qc_median_intensities(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = normalised_intensity_log2
) + ggplot2::ggtitle("After Normalization")

# Plot the two plots side-by-side

cowplot::plot_grid(
    b1, b2,
    ncol = 2
)

```

```{r}
# Imputation of Missing Values

# Impute missing values
protein_data_long <- impute_with_downshifted_normal(
    data = protein_data_long,
    intensity_log2 = normalised_intensity_log2,
    prctl = 0.05,
    downshift_mag = 1.5,
    downshift_min = 0.1
)

p1 <- qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = normalised_intensity_log2,
    plot_style = "violin"
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(
        angle = 90,
        hjust = 1
    )
) + ggplot2::ggtitle(
    "Non-Imputed Intensity Distribution"
)


p2 <- qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = imputed_intensity_log2,
    plot_style = "violin"
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(
        angle = 90,
        hjust = 1
    )
) + ggplot2::ggtitle(
    "Imputed Intensity Distribution"
)

cowplot::plot_grid(
    p1, p2,
    ncol = 2,
    align = "v"
)
```

```{r}
## Statistical Testing with Weighted Limma package

# Transform the long to wide format
# columns:Sample, rows: Protein, values: imputed_intensity_log2
protein_data_wide <- protein_data_long %>%
    dplyr::select(
        Sample,
        Protein,
        imputed_intensity_log2
    ) %>%
    tidyr::pivot_wider(
        names_from = Sample,
        values_from = imputed_intensity_log2
    ) %>%
    tibble::column_to_rownames(
        var = "Protein"
    )
View(protein_data_wide)

write.csv(protein_data_wide, "output/EIF2A_proteomics_log2_norm_impute.csv")
protein_data_wide <- read.csv("output/EIF2A_proteomics_log2_norm_impute.csv", row.names = 1)

protein_data_wide_non_imputed <- protein_data_long %>%
    dplyr::select(
        Sample,
        Protein,
        normalised_intensity_log2
    ) %>%
    tidyr::pivot_wider(
        names_from = Sample,
        values_from = normalised_intensity_log2
    ) %>%
    tibble::column_to_rownames(
        var = "Protein"
    )

write.csv(protein_data_wide_non_imputed, "output/EIF2A_proteomics_log2_norm_nonimpute.csv")
protein_data_wide_non_imputed <- read.csv("output/EIF2A_proteomics_log2_norm_nonimpute.csv", row.names = 1)


# Find the indices of the missing values
na_index <- which(is.na(protein_data_wide_non_imputed))

# Initialize the weight matrix
weight_matrix <- matrix(
    data = 1,
    nrow = nrow(protein_data_wide),
    ncol = ncol(protein_data_wide)
)
# Weighting of Missing Values
na_weight <- 0.0001
# Replace the missing values with the na_weight
weight_matrix[na_index] <- na_weight

View(weight_matrix)

# Create the design matrix

group_levels <- c("DMSO_ctrl", "DMSO_EIF2A", "Tg_ctrl", "Tg_EIF2A")
f <- factor(metadata$Group, levels = group_levels)
design <- model.matrix(~0+f)
colnames(design) <- group_levels
View(design) 

# Variables
pval_thr <- 0.05
log2_fc_thr <- 0

# Fit a linear model with weights
fit <- limma::lmFit(
    protein_data_wide,
    design = design,
    weights = weight_matrix
)


vs <- c(paste0(group_levels[2],"-",group_levels[1]),  
        paste0(group_levels[4],"-",group_levels[3]),  
        paste0(group_levels[3],"-",group_levels[1]), 
        paste0(group_levels[4],"-",group_levels[2]))

contrast_matrix <- limma::makeContrasts(vs[1],vs[2], vs[3], vs[4], levels = design)

fit2 <- limma::contrasts.fit(fit, contrast_matrix)
fit2 <- limma::eBayes(fit2)

library(openxlsx)
library(org.Hs.eg.db)
# Create a new Excel workbook
wb <- createWorkbook()
for (i in 1:length(vs)){
  df <- limma::topTable(fit2, n=Inf, coef=i, adjust="BH") %>%
    rownames_to_column(var = "symbol")
  
  
  df$entrez1 <- mapIds(org.Hs.eg.db, keys=c(df$symbol), column="ENTREZID", keytype="SYMBOL", multiVals="first") %>%
    as.character()
  df$entrez2 <- mapIds(org.Hs.eg.db, keys=c(df$symbol), column="ENTREZID", keytype="ALIAS", multiVals="first") %>% 
    as.character()
  df <- df %>% 
    mutate(entrez = coalesce(entrez1, entrez2), .keep = "unused")
  
  df$ENSG1 <- mapIds(org.Hs.eg.db, keys=c(df$symbol), column="ENSEMBL", keytype="SYMBOL", multiVals="first") %>%
    as.character()
  df$ENSG2 <- mapIds(org.Hs.eg.db, keys=c(df$symbol), column="ENSEMBL", keytype="ALIAS", multiVals="first") %>% 
    as.character()
  df <- df %>% 
    mutate(ENSG = coalesce(ENSG1, ENSG2), .keep = "unused")
  df$fullname <- mapIds(org.Hs.eg.db, keys=c(df$symbol), column="GENENAME", keytype="SYMBOL", multiVals="first") %>%
    as.character()
  df[df=="NULL"] <- NA

  #save to workbook
  addWorksheet(wb, vs[i])
  writeData(wb, vs[i], df)
}
saveWorkbook(wb, "output/EIF2A_proteomics_limma_DE_results.xlsx", overwrite = TRUE)
View(df)

```


```{r filter reliably detected proteins N>=2}

dmso.e_dmso.c <- read_excel("output/EIF2A_proteomics_limma_DE_results.xlsx", sheet = 1)
tg.e_tg.c <- read_excel("output/EIF2A_proteomics_limma_DE_results.xlsx", sheet = 2)
tg.c_dmso.c <- read_excel("output/EIF2A_proteomics_limma_DE_results.xlsx", sheet = 3)
tg.e_dmso.e <- read_excel("output/EIF2A_proteomics_limma_DE_results.xlsx", sheet = 4)

#protein_data_wide <- read.csv("output/EIF2A_proteomics_log2_norm_impute.csv", row.names = 1)
protein_data_wide_non_imputed <- read.csv("output/EIF2A_proteomics_log2_norm_nonimpute.csv", row.names = 1)

View(dmso.e_dmso.c)
View(protein_data_wide_non_imputed)
df = protein_data_wide_non_imputed
group_levels = c("DMSO.ctrl", "DMSO.EIF2A", "Tg.ctrl", "Tg.EIF2A")
samples = colnames(protein_data_wide_non_imputed)
vs <- c(paste0(group_levels[2],"_",group_levels[1]),  
        paste0(group_levels[4],"_",group_levels[3]),  
        paste0(group_levels[3],"_",group_levels[1]), 
        paste0(group_levels[4],"_",group_levels[2]))

reliable.list <- list()
reliable.list[[group_levels[1]]] <- rownames(df[rowSums(is.na(df[,grep(group_levels[1],samples)]))<2,])
reliable.list[[group_levels[2]]]  <- rownames(df[rowSums(is.na(df[,grep(group_levels[2],samples)]))<2,])
reliable.list[[group_levels[3]]]  <- rownames(df[rowSums(is.na(df[,grep(group_levels[3],samples)]))<2,])
reliable.list[[group_levels[4]]]  <- rownames(df[rowSums(is.na(df[,grep(group_levels[4],samples)]))<2,])
View(reliable.list)


# kept the proteins that have n >=2 per group for each comparison
dmso.e_dmso.c_reliable <- filter(dmso.e_dmso.c, symbol %in% intersect(reliable.list[[1]], reliable.list[[2]]) )

tg.e_tg.c_reliable <- filter(tg.e_tg.c, symbol %in% intersect(reliable.list[[3]], reliable.list[[4]]) )

tg.c_dmso.c_reliable <- filter(tg.c_dmso.c, symbol %in% intersect(reliable.list[[1]], reliable.list[[3]]) )

tg.e_dmso.e_reliable <- filter(tg.e_dmso.e, symbol %in% intersect(reliable.list[[2]], reliable.list[[4]]) )
View(dmso.e_dmso.c_reliable)

# Create a new Excel workbook
wb <- createWorkbook()
#save to workbook
addWorksheet(wb, vs[1])
writeData(wb, vs[1], dmso.e_dmso.c_reliable)
addWorksheet(wb, vs[2])
writeData(wb, vs[2], tg.e_tg.c_reliable)
addWorksheet(wb, vs[3])
writeData(wb, vs[3], tg.c_dmso.c_reliable)
addWorksheet(wb, vs[4])
writeData(wb, vs[4], tg.e_dmso.e_reliable)


saveWorkbook(wb, "output/EIF2A_proteomics_limma_DE_results_filtered.N2.xlsx", overwrite = FALSE)


```





```{r}
ggplot2::ggplot(
        stat_res_df,
        ggplot2::aes(
            x = log2_fc,
            y = -log10(adj_pvalues),
            color = significance,
            alpha = significance
        )
    ) +
    ggplot2::geom_point(size = 2) +
    ggplot2::geom_vline(
        xintercept = log2_fc_thr,
        linetype = "dashed",
        color = "darkgrey"
    ) +
    ggplot2::geom_vline(
        xintercept = -log2_fc_thr,
        linetype = "dashed",
        color = "darkgrey"
    ) +
    ggplot2::geom_hline(
        yintercept = -log10(pval_thr),
        linetype = "dashed",
        color = "darkgrey"
    ) +
    ggplot2::scale_color_manual(
        values = c(
            "Up regulated" = "#e63946",
            "Down regulated" = "#1d3557",
            "no significance" = "#b1a7a6"
        )
    ) +
    ggplot2::scale_alpha_manual(
        values = c(
            "Up regulated" = 1.0,
            "Down regulated" = 1.0,
            "no significance" = 0.2
        )
    ) +
    ggplot2::ggtitle("") +
    ggplot2::labs(
        x = "log2(Fold-Change)",
        y = "-log10(adjusted p-value)"
    )
```
```{r}
# Plot MA
ggplot2::ggplot(
        stat_res_df,
        ggplot2::aes(
            x = average,
            y = log2_fc,
            color = significance,
            alpha = significance
        )
    ) +
    ggplot2::geom_point(
        size = 3.5,
    ) +
    ggplot2::geom_hline(
        yintercept = log2_fc_thr,
        linetype = "dashed",
        color = "darkgrey"
    ) +
    ggplot2::geom_hline(
        yintercept = -log2_fc_thr,
        linetype = "dashed",
        color = "darkgrey"
    ) +
    # Draw a vertical line at the average intensity of 8
    ggplot2::geom_vline(
        xintercept = 8,
        linetype = "dashed",
        color = "#201717"
    ) +
    ggplot2::scale_color_manual(
        values = c(
            "Up regulated" = "#e63946",
            "Down regulated" = "#1d3557",
            "no significance" = "#b1a7a6"
        )
    ) +
    ggplot2::scale_alpha_manual(
        values = c(
            "Up regulated" = 1.0,
            "Down regulated" = 1.0,
            "no significance" = 0.2
        )
    ) +
    ggplot2::ggtitle("") +
    ggplot2::labs(
        x = "Average Intensity",
        y = "log2(Fold-Change)"
    )
```

# Enrichment Analysis with gProfiler2 package

# Get the list of proteins
# Background is the list of all proteins
background <- rownames(protein_data_wide)
# Initilize the query list
go_list <- list()
# Add the list of up regulated proteins
go_list[["up_regulated"]] <- rownames(
    stat_res_df[
        which(
            stat_res_df$significance == "Up regulated"
        ),
    ]
)
# Add the list of down regulated proteins
go_list[["down_regulated"]] <- rownames(
    stat_res_df[
        which(
            stat_res_df$significance == "Down regulated"
        ),
    ]
)

# Run the enrichment analysis with gProfiler2
gostres <- gprofiler2::gost(
    query = go_list[["up_regulated"]],
    organism = "hsapiens",  # Human
    user_threshold = 0.05,  # Enrichment p-value threshold
    custom_bg = background, # Custom Background
    domain_scope = "custom_annotated",
    sources = c(            # Source of annotations
        "GO:BP",
        "GO:MF",
        "GO:CC",
        "KEGG",
        "REAC"
        # "WP",
        # "HPA",
        # "CORUM",
        # "MIRNA",
        # "TF",
        # "HP"
    ),
    multi_query = FALSE,
    correction_method = "fdr",
)
# Calculate Gene Ratio
gostres$result$GeneRatio <- (
    gostres$result$intersection_size / gostres$result$term_size
)

gprofiler2::gostplot(
    gostres,
    capped = TRUE,
    interactive = TRUE,
```

